---
title: "Actors Investigation"
author: "Henry Savich"
output: github_document
---

In this notebook we are going to look at data from the IMDB movie dataset sourced from kaggle to determine if there is a relationship between actors/actresses and movie ratings

```{r}
suppressPackageStartupMessages(library(tidyverse)) #For convenient data frame manipulation
suppressPackageStartupMessages(library(ggplot2)) #For visualizing the data
```

IMDb title_principals lists every occurrence of an actor/actress in a movie

```{r}
tp = read_csv("data/IMDb\ title_principals.csv") %>%
  #the dataframe also includes writers, directors, cinmetographers etc., so
  #we'll filter to only get actors/actresses
  filter(category == "actor" | category == "actress") %>%
  select(imdb_title_id, imdb_name_id, category, characters)
```

```{r}
length(unique(tp$imdb_name_id))
```

Our dataset contains 132,123 actors / actresses. However, since we only want to look at English movies, we will need to add that information from another csv in order to filter.

```{r}
mv = read_csv("data/IMDb\ movies.csv") %>%
  select(imdb_title_id, language)

tp2 = left_join(tp, mv, by = "imdb_title_id") %>%
  filter(str_detect(language, "English"))

length(unique(tp2$imdb_name_id))
```

So out of the 132,123 actors/actresses in the database, 74,663 have acted in English language movies (56.5% of actors/actresses).

Let's now look at how many movies these actors/actresses were in.

```{r}
freq_df = count(tp2, imdb_name_id)

max(freq_df$n)
mean(freq_df$n)
median(freq_df$n)

ggplot(freq_df, aes(x = n)) +
  geom_histogram(bins = 100)
```

From this we learn:

-The actor/actress with the most English-language titles on IMDB has 132 (John Wayne)
-The mean number of English-language titles is 2.58
-The median number of titles is 1, which means most of our actors/actresses only appear in one title

Because we want to see distributions of ratings of movies for an individual actor, it doesn't make sense to try and gather information from an actor with only 1 movie, and for a small enough number of movies, the distribution of ratings won't really reflect the actor/actress, so we want to establish some cutoff number of movies. Let's look at how many actors/actresses we have that meet a given threshold:

```{r}

threshold = 2:100
num_actors = rep(0, 99)
for(i in threshold){
  num_actors[i-1] = sum(freq_df$n >= i)
}

ggplot(data.frame(threshold = threshold, num_actors = num_actors), aes(x = threshold, y = num_actors)) +
  geom_line() +
  geom_hline(yintercept = 1000, color = "red")

sum(freq_df$n >= 20)
```

If we decide we want to have at least 1,000 actors/actresses to look at, the threshold can be at most 23 films. 20 seems like a good round number to use as the threshold. This will leave us with 1424 people.

```{r}
tp3 = left_join(tp2, freq_df, by = "imdb_name_id") %>%
  filter(n >= 20)
```

Now we add in the actual names from IMDb names.csv and our actors/actresses dataframe is complete!

```{r}
id_to_names = read_csv("data/IMDb\ names.csv") %>%
  select(imdb_name_id, name)

people = left_join(tp3, id_to_names, by = "imdb_name_id")
```

# Actors Correlations with Ratings

The meat of this investigation will involve looking at how different actors effect the distribution of ratings of a movie.

The IMDb rating scale is 1-10, and we have a choice of how we want to look at rating. We'll start by using the avg_vote column in IMDb movies.csv

ˆ
```{r}
people = read_csv("data/IMDb\ movies.csv") %>%
  select(imdb_title_id, avg_vote, votes) %>%
  right_join(people, by = "imdb_title_id")
```

We could run into trouble here if some movies have a very small number of votes, as that would mean avg_vote would be very variable from a real representation of audience response.
Let's look at how num_votes is distributed:

```{r}
english_mv = read_csv("data/IMDb\ movies.csv") %>%
  filter(str_detect(language, "English"))

num_votes = english_mv$votes

sprintf("There are %d English-language movies", length(num_votes))
sprintf("The max number of votes is %d.", max(num_votes))
sprintf("The min number of votes is %d.", min(num_votes))
sprintf("The mean number of votes is %f.", mean(num_votes))
sprintf("The median number of votes is %d.", median(num_votes))

l100 = sum(num_votes < 100)
sprintf("There are %d movies with less than 100 votes, which is %f of movies",
        l100, l100/length(num_votes))

l1000 = sum(num_votes < 1000)
sprintf("There are %d movies with less than 1000 votes, which is %f of movies",
        l1000, l1000/length(num_votes))

hist(num_votes)
```
The distribution is so right skewed, that with the default parameters, the histogram doesn't tell us much.

It looks like only movies with 99 or more ratings are included in the dataset, so we can go ahead without filtering movies with too few ratings.

```{r}
actor_ratings = people %>%
  group_by(imdb_name_id) %>%
  summarize(avg_rating = mean(avg_vote), vote_var = var(avg_vote)) %>%
  left_join(id_to_names, by = "imdb_name_id") %>%
  left_join(freq_df, by = "imdb_name_id")

hist(actor_ratings$avg_rating)
```

We get to see some interesting stats now by looking at the actor_ratings table!

The actor with the worst average rating is Joe Estevez (Charlie Sheen's uncle), who is credited in 23 movies with an average rating of 3.92.

The actor with the highest average rating is Leonardo DiCaprio,who is credited in 27 movies with an average rating of 7.36

```{r}
sprintf("The average average rating is %f", mean(actor_ratings$avg_rating))
sprintf("The median average rating is %f", median(actor_ratings$avg_rating))
```

The average and median are pretty close, but the distribution is definitely slightly left-skewed (by people like Joe Estevez).

To see how strong the correlation between actor and avg_rating is, let's do a simulation of if each actor just starred in as many random movies.

```{r}
avg_ratings = actor_ratings$avg_rating
n = actor_ratings$n
sim_avg_ratings = rep(0, length(n))
movie_ratings = english_mv$avg_vote
for(i in 1:length(n)){
  sim_avg_ratings[i] = mean(sample(movie_ratings, n[i]))
}

hist(sim_avg_ratings)
```

Let's compare this distribution to the actual distribution of avg_ratings:

```{r}
compare_sim = data.frame(actual = actor_ratings$avg_rating, sim = sim_avg_ratings) %>%
  pivot_longer(c(actual, sim), names_to = "type")

ggplot(compare_sim, aes(x=value, fill=type)) +
    geom_histogram( color="#e9ecef", alpha=0.6, position = 'identity') +
    scale_fill_manual(values=c("#12C734", "#050399"))
```

We can see the distributions look very different! We can conclude there is a strong correlation between actors and movie_ratings.
