---
title: "budget_investigation"
author: "Cam Rondeau"
date: "11/28/2021"
output: html_document
---

The code below investigates the relationship between production budgets and movie ratings. The data I will be using to conduct this investigation is the IMDb movie database, which was downloaded via Kaggle.

```{r include=FALSE}
library(dplyr)
library(stringr)
library(ggplot2)
library(priceR)
```


I will first load in the movie data. For this investigation, I will only be looking at movies made in the United States. I can also only analyze the movies that actually have data on their budget. 

```{r}
movies <- read_csv("IMDb_movies.csv") %>%
  filter(country == "USA" & budget != "")
```

This gives us 10,488 movies that have data on their budget. Next, I need to do some data cleaning in order to be able to analyze the budgets. The budget data in the raw excel starts with a currency sign, so I must first filter for USD, then get rid of the $ and convert the string to numeric. I will also select the columns I need and convert to numeric types where needed.

```{r}
budgets <- movies %>%
  filter(str_detect(budget, "\\$")) %>%
  mutate(budget = as.numeric(gsub("\\$", "", budget)), duration = as.numeric(duration)) %>%
  select(original_title, year, budget, avg_vote, duration, worlwide_gross_income) %>%
  arrange(desc(budget))
```


When looking at historical data on money, its important to consider inflation ad how that may effect the perception of certain dollar amounts from a long time ago. A massive budget in 1950 may seem like a tiny budget in 2020, so we need to adjust these amounts for inflation in order to get an accurate representation of how budget may effect movie ratings.

```{r}
mean(budgets$year)
```
The mean year for the movie data is 1991, so we definitely have data on a lot of old movies that need to have adjusted budgets.

```{r}
budgets <- budgets %>%
  mutate(adj_budget = adjust_for_inflation(budget, from_date=as.numeric(year), to_date=2020, country = "US")) %>%
  arrange(desc(adj_budget))
```

```{r}
ggplot(budgets, aes(adj_budget, avg_vote)) + 
  geom_point() + 
  geom_smooth(method="lm", se=FALSE) +
  labs(title = "Relationship between Budget and Movie Rating",
       x="Inflation Adjusted Budget",
       y="Rating") +
  scale_x_continuous(labels=scales::dollar_format()) +
  theme_bw()

cor(budgets$adj_budget, budgets$avg_vote)
```

As you can see from the graph, there is a positive relationship between adjusted budget and movie rating. The correlation coefficient is 0.275, indicating that there is a positive relationship, albeit not an overly strong one. There is an overabundance of movies with very small budgets. As the budget increases, there are fewer movies, but the rating for these higher budget movies is skewed to higher ratings. At the high end of movie budgets (>$300,000,000), there aren't any movies rated below 6.0.

Using the same scatter plot as above, it will also be helpful to group movies based on their budget amounts, and looking at the average rating of each budget group. From looking at the plot above, I will consider low budget movies to be <$50M, medium budgets to be between $50M and $125M, higher budget movies to be between $125M and $225M, and blockbuster type movies to be >$225M.

```{r}
budgets <- budgets %>%
  mutate(budget_group = case_when(
    adj_budget <= 50000000 ~ "<$50M",
    adj_budget > 50000000 & adj_budget <= 125000000 ~ "$50M - $125M",
    adj_budget > 125000000 & adj_budget < 225000000 ~ "$125M - $225M",
    adj_budget >= 225000000 ~ ">$225M",
  )) 

group_means <- budgets %>% 
  group_by(budget_group) %>% 
  summarise(mean = mean(avg_vote)) %>% 
  arrange(mean)

low <- as.numeric(group_means[1,2])
med <- as.numeric(group_means[2,2])
high <- as.numeric(group_means[3,2])
block <- as.numeric(group_means[4,2])


ggplot(budgets, aes(adj_budget, avg_vote, col=budget_group)) + 
  geom_point(alpha=0.5) +
  geom_hline(yintercept=group_means$mean, size=1, col=c("mediumorchid2", "cyan3", "olivedrab3", "coral2")) +
  scale_color_discrete(name="Budget Group", limits = c(">$225M", "$125M - $225M", "$50M - $125M", "<$50M")) +
  labs(title = "Average Movie Rating by Budget Group",
       x="Inflation Adjusted Budget",
       y="Rating") +
  scale_x_continuous(labels=scales::dollar_format()) +
  theme_bw()
  
```

As you can see in the plot above, each successive budget group has a higher average movie rating. 

```{r}
group_means
```


Another thing that should be considered when looking at movies budgets is the movie runtime. Very long movies have more scenes and cost more money to make by default, whereas very short movies require much less money to make. When looking at movie ratings, the movie duration does determine whether it receives a high rating or not. Because of this, we should also standardize the movie budgets to account for duration. So, I will add a new variable that takes the adjusted budget over the movie duration, and investigate how this feature effects movie ratings.

```{r}
budgets <- budgets %>%
  mutate(adj_budget_duration = adj_budget / duration) %>%
  arrange(desc(adj_budget_duration))

ggplot(budgets, aes(adj_budget_duration, avg_vote)) + 
  geom_point() + 
  geom_smooth(method="lm", se=FALSE) +
  labs(title = "Relationship between Duration Adjusted Budget and Movie Rating",
       x="Inflation Adjusted Budget over Duration",
       y="Rating") +
  scale_x_continuous(labels=scales::dollar_format()) +
  theme_bw()

cor(budgets$adj_budget_duration, budgets$avg_vote)
```


The relationship between movie budget and movie rating is actually slightly less after adjusting the budget to account for movie duration. There could be two reasons for this. One is that short films with low movie ratings now have much higher adjusted budgets, which caused the regression line to get pulled down to the right. The second reason may be that very long movies tend to have higher ratings, so their new adjusted budget is less than before, which pulls the left side of the regression line up. Both scenarios lead to a slightly flatter regression line, but even still, there is still a positive relationship between budget and movie rating.


The last thing I will investigate is how box office performance relates to high movie ratings. Before, we looked at movie budgets, which, intuitively, has an actual effect on the popularity and quality of a movie. Now, I want to investigate the correlation between revenue and movie ratings. To do so, I need to perform similar data cleaning and filtering as before. I will also adjust the revene numbers for inflation.

```{r}
revenue <- budgets %>%
  filter(worlwide_gross_income!="") %>%
  mutate(worlwide_gross_income = as.numeric(gsub("\\$", "", worlwide_gross_income))) %>%
  mutate(adj_revenue = adjust_for_inflation(worlwide_gross_income, from_date=as.numeric(year), to_date=2020, country = "US")) %>%
  arrange(desc(adj_revenue))
```

Now I will plot the relationship between box office revenue earnings and movie rating.

```{r}
ggplot(revenue, aes(adj_revenue, avg_vote)) + 
  geom_point() + 
  geom_smooth(method="lm", se=FALSE) +
  labs(title = "Relationship between Revenue and Movie Rating",
       x="Inflation Adjusted Revenue Earnings",
       y="Rating") +
  scale_x_continuous(labels=scales::dollar_format()) +
  theme_bw()

cor(revenue$adj_revenue, revenue$avg_vote)
```

As you can see, there are much fewer movies that we have both budget data and revenue data for. However, for the movies we do have both data points for, the correlation between revenue and movie ratings is almost identical to the relationship between budgets and movie ratings. There are again many movies with very small revenue figures, and the movies with revenue >$200M all have movie ratings over 7.0.


Overall, it is clear that movie ratings increase as movie budgets increase. There are some limitations with the data that may skew the results a bit here, particularly how there are so many movies with very low budgets, but there is no doubt a positive correlation. More money doesn't always mean a better movie, but the highest budget movies have, on average, the highest ratings.













